Notes on Control Plane Inference

### 2021-07-21 inference.tex
fix typo in instrumentation def
add rewrite rules

@@

% Theorems
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{conjecture}{Conjecture}
\newtheorem{corollary}{Corollary}

@@

\begin{figure}[htp]
\[\begin{array}{r >{\triangleq}cl}
  \multicolumn{3}{l}{(\cdot)\instr : \Cmd \to \Cmd} \\
  (h.f := e)\instr && h.f := e \\
  (\assume b)\instr && \assume b \\
  (\assert b)\instr && \assert b\\
  (c_1;c_2)\instr && c_1\instr;c_2\instr\\
  (c_1 \choice c_2)\instr && c_1\instr \choice c_2\instr\\
  (t.\apply())\instr &&\\
  \multicolumn{3}{l}{
    \qquad
    \begin{array}[t]{l} \displaystyle
      \assume t.\matchrow(\rho_t); \\
      \assume \gamma_t.\keys = t.\keys; \\
    \displaystyle \bigchoice_{i=0}^{|t.\action| -1}
    \left(\begin{array}{l}
      \assume \gamma_t.\miss = 0;\\
      \assume \rho_t.\action = i;\\
      \assume \gamma_t.\hitAction = \rho_t.\id;\\
      t.\action[i](\rho_t.\data)
    \end{array}\right)\\
    \choice (\assume \gamma_t.\miss = 1; \SKIP)  \\
  \end{array}}
  \end{array}
  \]
  \caption{Symbolic Table Instrumentation. For simplicity, we
    assume each action in a table has the same number (and types) of parameters}
  \label{fig:table-instrument}
\end{figure}

@@

>>>>

%% inference rules
\newcommand{\andalso}{\quad}
\newcommand{\pad}[1]{\mbox{}\;#1\;\mbox{}}
\newcommand{\infrule}[3]{\ensuremath{\frac{\pad{#1}}{\pad{#2}}\;{\textsc{#3}}}}

@@

% Theorems
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{conjecture}{Conjecture}
\newtheorem{corollary}{Corollary}
\newtheorem{proposition}{Proposition}

@@

\begin{figure}[htp]
\[\begin{array}{r >{\triangleq}cl}
  \multicolumn{3}{l}{(\cdot)\instr : \Cmd \to \Cmd} \\
  (h.f := e)\instr && h.f := e \\
  (\assume b)\instr && \assume b \\
  (\assert b)\instr && \assert b\\
  (c_1;c_2)\instr && c_1\instr;c_2\instr\\
  (c_1 \choice c_2)\instr && c_1\instr \choice c_2\instr\\
  (t.\apply())\instr &&\\
  \multicolumn{3}{l}{
    \qquad
    \begin{array}[t]{l} \displaystyle
      \assume \gamma_t.\keys = t.\keys; \\
    \displaystyle \bigchoice_{i=0}^{|t.\action| -1}
    \left(\begin{array}{l}
      \assume t.\matchrow(\rho_t); \\
      \assume \gamma_t.\miss = 0;\\
      \assume \rho_t.\action = i;\\
      \assume \gamma_t.\hitAction = \rho_t.\id;\\
      t.\action[i](\rho_t.\data)
    \end{array}\right)\\
    \choice (\assume \gamma_t.\miss = 1; \SKIP)  \\
  \end{array}}
  \end{array}
  \]
  \caption{Symbolic Table Instrumentation. For simplicity, we
    assume each action in a table has the same number (and types) of parameters}
  \label{fig:table-instrument}
\end{figure}

@@

\subsection{Heuristic Rules}

Here we posit a collection of heuristic elimination rules for quantified
variables that seem to be effective (at least in some small examples). The goal
here is to define a collection of rules that will remove quantifiers for a broad
class of programs.

\begin{figure}[H]
  \[
    \begin{array}{>{\displaystyle}c}
      \infrule{
      \psi[\forall x. \varphi_{1} \Rightarrow \varphi_{2}] \andalso x \not \in \fvs(\varphi_{1})}{\psi[\varphi_{1} \Rightarrow \forall x. \varphi_{2}]}{ReduceScopeR} \qquad

      \infrule{\psi[\forall x. \varphi_{1} \Rightarrow \varphi_{2}] \andalso x \not \in \fvs(\varphi_{2})}{\psi[(\forall x. \neg \varphi_{1}) \vee \varphi_{2}]}{ReduceScopeL} \\[1em]

      \infrule{\psi[\forall x. \varphi] \andalso x \not \in \fvs(\varphi)}{\psi[\varphi]}{UnusedUVar}
      \qquad
      \infrule{\psi[\exists x. \varphi] \andalso x \not \in \fvs(\varphi)}{\psi[\varphi]}{UnusedEVar}  \\[1em]

      \infrule{\psi[\forall x. x = e]}{\FALSE}{UnivEq}
      \qquad
      \infrule{\psi[\exists x. x = e]}{\TRUE}{ExtEq}
    \end{array}
  \]
\end{figure}

Now we prove that these rules are sound, that is, applying them to a formulae
$\psi$ produces a logically equivalent formula $\psi'$.

\begin{proposition}[Rewrite Soundness]
  The above rules are sound.
\end{proposition}
\begin{proof}
  Consider each rule separately.
  \begin{enumerate}[align=left]
    \item[(\textsc{ReduceScopeR})]
    Let $x$, $\varphi_{1}$ and $\varphi_{2}$ be given such that $x \not \in \fvs(\varphi)$. Let $\mu$ be a (total) model. Show that
    \[\mu \models \forall x. \varphi_{1} \Rightarrow \varphi_{2} \iff \mu \models \varphi_{1} \Rightarrow \forall x. \varphi_{2}\]
    Show each direction separately.
    \begin{enumerate}[align=left]
      \item[($\Rightarrow$)] Assume $\mu \models \forall x. \varphi_{1} \Rightarrow \varphi_{2}$. Further, assume $\mu \models \varphi_1$. Let $v$ be given to show $\mu[x \mapsto v] \models \varphi_{2}$. By the semantics of universal quantification $\mu[x \mapsto v] \models \varphi_{1} \Rightarrow \varphi_{2}$. Since $x$ does not occur in $\varphi_{1}$, and $\mu \models \varphi_{1}$ we know $\mu[x \mapsto v] \models \varphi_{1}$. By definition of the semantics of implication, we know $\mu[x \mapsto v] \models \varphi_{2}$. $\checkmark$

      \item[$(\Leftarrow)$] Assume $\mu \models \varphi_1 \Rightarrow \forall x. \varphi_{2}$.
      Let $v$ be given to show $\mu[x \mapsto v] \models \varphi_{1} \Rightarrow \varphi_{2}$.
      Assume $\mu[x \mapsto v] \models \varphi_{1}$ (otherwise we'd be done), to show $\mu[x \mapsto v] \models \varphi_{2}$.
      This assumption gives us $\mu \models \varphi_{1}$ since $x \not \in \fvs(\varphi_{1})$. Then
      $\mu \models \forall x. \varphi_{2}$, which means, by definition of the semantics of universal quantification, that $\mu[x \mapsto v] \models \varphi_{2}$. $\checkmark$
    \end{enumerate}

    \item[(\textsc{ReduceScopeL})]
    Let $x$, $\varphi_{2}$, and $\varphi_2$ such that $x \not \in \fvs(\varphi_{2})$.
    Let $\mu$ be a total model, to show that
    \[\mu \models \forall x. \varphi_{1} \Rightarrow \varphi_{2}
      \iff
      \mu \models (\forall x. \neg \varphi_{1}) \vee \varphi_{2}
    \]
    Prove each direction separately:
    \begin{enumerate}[align=left]
      \item[($\Rightarrow$)]
      Assume $\mu \models \forall x. \varphi_{1} \Rightarrow \varphi_{2}$.
      Assume $\mu \models \exists x. \varphi_{1}$ to show $\mu \models \varphi_{2}$.
      Let $v$ be such that $\mu[x \mapsto v] \models \varphi_{1}$.
      Then we know that $\mu[x \mapsto v] \models \varphi_{1} \Rightarrow \varphi_{2}$.
      Further, since $\mu[x \mapsto v] \models \varphi_{1}$, we know $\mu[x \mapsto v] \models \varphi_{2}$. Since $x \not \in \fvs(\varphi_{2})$, conclude that $\mu \models \varphi_{2}$.
      \hfill $\checkmark$

      \item[($\Leftarrow$)]
      Assume $\mu \models (\forall x. \neg \varphi_{1}) \vee \varphi_{2}$. To show that $\mu \models \forall x. \varphi_{1} \Rightarrow \varphi_{2}$. Let $v$ be arbitrary, and assume that $\mu[x \mapsto v] \models \varphi_{1}$ (otherwise we'd be done), to show that $\mu[x \mapsto v] \models \varphi_{2}$.

      Our assumption gives us two cases. If $\mu \models \varphi_{2}$ then the result follows since $x \not \in \fvs(\varphi_{2})$ means $\mu[x \mapsto v] \models \varphi_{2}$.

      In the second case, $\mu \models (\forall x. \neg \varphi_{1})$, which means that for our $v$,
      $\mu[x \mapsto v] \not\models \varphi_{1}$. This contradicts the assumption that $\mu[x \mapsto v] \models \varphi_{1}$. And we're done.

    \end{enumerate}

    \item[(\textsc{UnusedUVar})] \todo[inline]{Prove this case}
    \item[(\textsc{UnusedEVar})] \todo[inline]{Prove this case}
    \item[(\textsc{UnivEq})] \todo[inline]{prove this case}
    \item[(\textsc{ExtEq})] \todo[inline]{prove this case}

  \end{enumerate}
